---
title: "NYPD Incidents Project Report"
author: "Benjamin Ahlbrecht"
date: "2/27/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Question of Interest
Here, we hope to clarify the nature of murder cases in New York. Is there an effect given the race, age, sex, or location of the individuals? Are murders on the rise or the decline? What possible sources of bias remain in the analysis?

### Library Preamble
```{r libraries, echo=TRUE}
library('tidyverse')
library('lubridate')
```


### Import NYPD Incident Data
The data is supplied by [data.gov](https://catalog.data.gov/dataset). The dataset used in this analysis is the NYPD Shooting Incident Data (Historic), which provides a breakdown of shooting incidents in NYC from 2006--2020 and can be found [here](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic)
```{r importing, echo=TRUE}
data.url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
data.raw <- read.csv(data.url)
```

### Clean Data: Remove Unnecessary Columns
```{r cleaning, echo=TRUE}
data.raw.cols <- colnames(data.raw)

data.raw.cols

# Determine the columns we don't need to keep
data.raw.cols.bad <- c(1, 5, 6, 7, 17, 18, 19)

# We'll change our naming style a bit
data.clean.cols <- c(
    "Date", "Time", "Borough", "Murder", "PerpAge", "PerpSex", "PerpRace",
    "VictimAge", "VictimSex", "VictimRace", "Xcoord", "Ycoord"
)

# Create cleaned data
data.full <- data.raw[-data.raw.cols.bad] %>%
    setNames(data.clean.cols) %>%
    arrange(Date, Time) %>%
    mutate(Date=mdy(Date)) %>%
    mutate(Time=hms(Time)) %>%
    arrange(Date, Time)

# Replace all empty cells with NA
data.full.dims <- dim(data.full)
data.full.nrows <- data.full.dims[1]
data.full.ncols <- data.full.dims[2]

# For some reason 2 values in age are 940 and 224?.. Set to NA as well
data.full.bad.values <- c("", "UNKNOWN", "U", "940", "224")
for (i in 1:data.full.nrows) {
    for (j in 1:data.full.ncols) {
        val <- data.full[i, j]
        if (val %in% data.full.bad.values) {
            data.full[i, j] = NA
        }
    }
}
```

### Data Summary
```{r summary, echo=FALSE}
# Make sure we don't have any funky values left
cat("\nUnique Values: Murder:\n", unique(data.full$Murder))
cat("\nUnique Values: PerpAge:\n", unique(data.full$PerpAge))
cat("\nUnique Values: PerpSex:\n", unique(data.full$PerpSex))
cat("\nUnique Values: PerpRace:\n", unique(data.full$PerpRace))
cat("\nUnique Values: VictimAge:\n", unique(data.full$VictimAge))
cat("\nUnique Values: VictimSex:\n", unique(data.full$VictimSex))
cat("\nUnique Values: VictimRace:\n", unique(data.full$VictimRace))
summary(data.full)
```

### Create Additional DataFrames for Analysis and Modeling
```{r refining, echo=TRUE}
# Select all murders and sort by date and time

data.murders <- data.full %>%
    filter(Murder == 'true')

# Count how many murders occurred each month
data.murders.by.month <- as.data.frame(
    table(
        data.murders %>%
        mutate(Year=year(Date), Month=month(Date)) %>%
        unite(YearMonth, Year, Month, sep="-") %>%
        mutate(YearMonth=ym(YearMonth)) %>%
        select(YearMonth)
    )
) %>%
    setNames(c("Date", "Murders")) %>%
    mutate(Date=ymd(Date))

# Count how many murders occured in each borough
data.murders.by.borough <- as.data.frame(
    table(data.murders %>% select(Borough))) %>%
    setNames(c("Borough", "Murders")
) %>%
    arrange(-Murders)

# Is there a correlation between perpetrator and victim age?
data.murders.by.age <- as.data.frame(
    table(data.murders %>% select(c(PerpAge, VictimAge)))
) %>%
    setNames(c("PerpAge", "VictimAge", "Murders"))


# Is there a correlation between perpetrator and victim race?
data.murders.by.race <- as.data.frame(
    table(data.murders %>% select(c(PerpRace, VictimRace)))
) %>%
    setNames(c("PerpRace", "VictimRace", "Murders"))


# Is there a correlation between perpetrator and victim sex?
data.murders.by.sex <- as.data.frame(
    table(data.murders %>% select(c(PerpSex, VictimSex)))
) %>%
    setNames(c("PerpSex", "VictimSex", "Murders"))
```


### Data Analysis and Modeling
```{r analysis_and_modeling, echo=TRUE}
# [1] Create a line plot to show the number of murders each month
data.murders.by.month.plot <- ggplot(data=data.murders.by.month) +
    aes(x=Date, y=Murders) +
    scale_x_date(date_labels="%Y", date_breaks="1 year") +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    geom_line() +
    theme_classic() +
    ggtitle("Monthly Murders in New York")

# It seems like we can fit a linear regression model relatively well
data.murders.by.month.mod <- data.murders.by.month.plot +
        geom_smooth(method="lm")

data.murders.by.month.plot
data.murders.by.month.mod

# [2] Create a simple bar chart to visualize murders by borough
data.murders.by.borough.plot <- ggplot(data=data.murders.by.borough) +
    aes(x=reorder(Borough, -Murders), y=Murders) +
    geom_col() +
    theme_classic() +
    ggtitle("Number of Murders in Each Burough") +
    xlab("Borough")

data.murders.by.borough.plot

# [3] Create a heatmap to visualize perpetrator Vs. victim age
data.murders.by.age.plot <- ggplot(data=data.murders.by.age) +
    aes(x=PerpAge, y=VictimAge, fill=Murders) +
    geom_tile() +
    scale_fill_distiller(palette="YlOrRd") +
    theme_classic() +
    geom_text(aes(label=Murders)) +
    ggtitle("Ages Involved in Murders")

data.murders.by.age.plot

# [4] Create a heatmap to visualize perpetrator Vs. victim race
data.murders.by.race.plot <- ggplot(data=data.murders.by.race) +
    aes(x=PerpRace, y=VictimRace, fill=Murders) +
    geom_tile() +
    scale_fill_distiller(palette="YlOrRd") +
    theme_classic() +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    geom_text(aes(label=Murders)) +
    ggtitle("Races Involved in Murders")

data.murders.by.race.plot

# [5] Create a heatmap to visualize perpetrator Vs. victim sex
data.murders.by.sex.plot <- ggplot(data=data.murders.by.sex) +
    aes(x=PerpSex, y=VictimSex, fill=Murders) +
    geom_tile() +
    scale_fill_distiller(palette="YlOrRd") +
    theme_classic() +
    geom_text(aes(label=Murders)) +
    ggtitle("Perp and Victim Genders in Murders")

data.murders.by.sex.plot
```

### Conclusion and Sources of Bias
The analysis here appears to highlight a number of key results:


1. The number of murders occurring in New York has been steadily decreasing each month since 2006, albeit some distinct outliers.

2. It appears that young to middle-aged (18--44) are more likely to both be a murder victim and a perpetrator. Younger and older individuals are less likely to be involved in a murder

3. Black-on-black, black-on-white hispanic and white hispanic-on-white hispanic are the most common races involved in murders in New York.

4. Males are much more likely to be the perpetrators than females with male-on-male murder being significantly more frequent than any other.

5. Murders occur most-frequently in Brooklyn and least-frequently in Staten Island.

It is important to acknowledge likely sources of bias from these results to avoid forming potentially erroneous conclusions. A likely mistake on all fronts is that the data is not conditioned on total population statistics. That is, simply because black-on-black murders are much more prominent than other combinations does not mean that other races are less likely to murder than black individuals. To assign causation without any further evidence would be ridiculous It may be, for instance, that a much higher percentage of the population in New York is black and, as such, we expect to see more black-on-black murders even if the probability any given race would murder another is uniform across all races.

Additionally, while young to middle-aged (18--44) are more likely to be involved in a murder, it does not mean that they are inherently more violenced than those who are younger or older. It may be the case that individuals aged 18--44 are simply more exposed to a harsh environment and are required to hard life decisions regarding how they live their lives. In other words, many poor individuals in New York are stuck between "a rock and a hard place": they must make hard decisions that promote a risky lifestyle. This risky lifestyle increases the probability that they will eventually become involved in a murder, either as the victim or the perpetrator. To that end, there is nothing to say here that individuals aged 18--44 are inherently more violent that any others.

All in all, this analysis is a first-step in examining the muddy world of murder incidents. There are many variables to consider and condition upon, and any suggestion of causation must be met with suspicion. Much more work would be needed with input from a number of domain experts to extract causation from a study such as this.